Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2018-11-16 mprop/mprop.tex
Making the proposal more readable and moving the derivatives out of the proposal

\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\diag}{diag}

@@

\newcommand{\dm}{\frac{\partial}{\partial\mathbf{m}}}
\newcommand{\dS}{\frac{\partial}{\partial\mathbf{S}}}
\newcommand{\da}{\frac{\partial}{\partial\alpha_j}}
\newcommand{\db}{\frac{\partial}{\partial\beta_j}}

\section{Introduction}
%briefly explain the context of the project problem
% TODO: introduce a data set
% TODO: example application for motivation (why would the world be better)
% TODO: introduce research questions

Inverse reinforcement learning (IRL)---a problem proposed by Russell in 1998
\cite{DBLP:conf/colt/Russell98}---asks us to find a reward function for a Markov
decision process that best explains a set of given demonstrations. IRL is
important because reward functions can be hard to define manually
\cite{DBLP:conf/icml/PieterN04,DBLP:journals/corr/abs-1806-06877}, and rewards
are not entirely specific to a given environment, allowing one to reuse the same
reward structure in previously unseen environments
\cite{DBLP:journals/corr/abs-1806-06877,DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
Moreover, IRL has seen a wide array of applications in autonomous vehicle
control
\cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16}
and learning to predict another agent's behaviour
\cite{DBLP:journals/ai/BogertD18,DBLP:conf/aaai/VogelRGR12,ziebart2008maximum,DBLP:conf/huc/ZiebartMDB08,DBLP:conf/iros/ZiebartRGMPBHDS09}.
Most approaches in the literature (see Section \ref{literature}) make a
convenient yet unjustified assumption that the reward function can be expressed
as a linear combination of features. One proven way to abandon this assumption
is by representing the reward function as a Gaussian process
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11,DBLP:journals/corr/abs-1208-2112}.

@@

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood estimates for $\bm\lambda$ and $\mathbf{u}$, and
made an assumption that $\mathbf{r}$ in \eqref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
variational inference to approximate the full posterior distribution
$\posterior$. \emph{Variational inference} (VI) is an approximation technique
for probability densities \cite{blei2017variational}. Let $\approximation$ be
our approximating family of probability distributions for $\posterior$ with its
own hyperparameter vector $\bm\nu$. Then the job of VI algorithms is to optimise
$\bm\nu$ in order to minimise the \emph{Kullback-Leibler} (KL) divergence
between the original probability distribution and our approximation.  KL
divergence (asymmetrically) measures how different the two distributions are,
and in this case can be defined as \cite{blei2017variational}
\[ \begin{split}
    \DKL(\approximation || \posterior) &= \Eq [\log\approximation -
    \log\posterior ] \\
    &= \Eq[\log\approximation - \log\pfull] \\
    &+ \Eq[\log p(\mathcal{D}, \mathbf{X_u})].
  \end{split}
\]
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as Lebesgue integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L}(\bm\nu) &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\bm\lambda\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\end{equation}

By considering full probability distributions instead of point estimates---as
long as the approximations are able to capture important features of the
posterior---our predictions are likely to be more accurate and rely on fewer
assumptions. Moreover, we hope to make use of various recent advancements in VI
for both time complexity and approximation distribution fit (see Section
\ref{literature}), making the resulting algorithm competitive both in terms of
running time and model fit.

@@

@@

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as neural networks)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, while
providing a time complexity of $\mathcal{O}(m^3)$. Below we provide a short
overview of various assumptions that have been used in approximating
\emph{sparse} GPs (i.e., GPs that use inducing points), following on a paper by
Qui\~{n}onero-Candela and Rasmussen \cite{DBLP:journals/jmlr/CandelaR05}.

@@

% TODO: expand on this
Table \ref{table:approximations} further summarises some of the recent and/or
influential GP approximation approaches that might be suitable for our GP IRL
model. In order to derive a reliable ELBO, the inducing points should be either
fixed or modelled by a probability distribution (none of the approaches do
that), while hyperparameters cannot be variational (the reason will be explained
in Section \ref{sec:proposed_approach}).

\begin{itemize}
\item We would also like to avoid having the hyperparameters fixed, as learning
  them efficiently would introduce a separate problem. As the approach by Hoang
  et al. does not seem to be easily extendable to modelled hyperparameters, it
  is unsuitable to our needs.
\item (TODO: this will have comments about other papers as well and will be
  restructured into a paragraph) The variational Fourier features (VFF)
  algorithm by Hensman et al. \cite{DBLP:journals/jmlr/HensmanDS17} is only
  defined for Mat\'ern kernels, which is likely to be too restrictive for our
  situation (the original paper on using GPs for IRL
  \cite{DBLP:conf/nips/LevinePK11} used the automatic relevance detection kernel
  that has weights controlling how important each feature is). While extending
  VFF to support a flexible class of kernels defined by Wilson and Adams
  \cite{pmlr-v28-wilson13} is an interesting and promising avenue of work, it is
  likely to be beyond the scope of this project.
\item The derivation of the ELBO in the work of Peng et al.
  \cite{DBLP:conf/icml/PengZZQ17} relies primarily on the fact that the evidence
  for a GP is a Gaussian, while in our case the evidence can be defined in
  several ways depending on the model, but must always involve $\mathcal{D}$,
  making the main idea of the paper inapplicable to IRL.
\end{itemize}

Since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing a wide range of possible probability
distributions. The primary way to represent complex posteriors in VI is by using
\emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Unfortunately, this
parametrisation also means that the gradient of the joint probability
distribution w.r.t. variational parameters $\nabla_{\bm\nu}\pfull$ is no
longer zero, making an analytic expression for ELBO impossible using the usual
methods. While it may be possible to have an approximating distribution for the
flow parameters, it is uncertain how such an algorithm would behave, as it
would have to perform optimisation in a space with significantly more dimensions.

@@

\subsection{The Structure of the Approximating Distribution}

@@

\subsection{Derivatives}
% TODO: some kind of intro

\subsubsection{$\partial/\partial\mathbf{m}$}

We begin by removing terms independent of $\mathbf{m}$:
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{m}} =
    &- \frac{1}{2}\dm[\mathbf{m}^\intercal\mathbb{E}[\Kuu^{-1}]\mathbf{m}]
    + \dm[\mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m}] \\
    &- \sum_{i=1}^N \sum_{t=1}^T \dm\mathbb{E}[V_r(s_{i,t})] -
      \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t},
      s')\dm\mathbb{E}[V_r(s')].
  \end{split}
\]
Here
\begin{align*}
  \dm[\mathbf{m}^\intercal\mathbb{E}[\Kuu^{-1}]\mathbf{m}] &= (\mathbb{E}[\Kuu^{-1}] + \mathbb{E}[\Kuu^{-1}]^\intercal)\mathbf{m}, \\
  \dm[\mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m}] &= \mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}],
\end{align*}
and
\begin{equation} \label{eq:1}
  \begin{split}
    \dm\mathbb{E}[V_r(s)] &= \dm\iiint V_r(s) p(\mathbf{r} | \bm\lambda,
    \mathbf{X_u}, \mathbf{u}) \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})
    q(\bm\lambda)\,d\mathbf{r}\,d\mathbf{u}\,d\bm\lambda \\
    &= \iiint V_r(s) p(\mathbf{r} | \bm\lambda,
    \mathbf{X_u}, \mathbf{u}) \dm\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})
    q(\bm\lambda)\,d\mathbf{r}\,d\mathbf{u}\,d\bm\lambda,
  \end{split}
\end{equation}
where
\[
  \begin{split}
    \dm\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S}) &=
    \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\dm\left[-\frac{1}{2}(\mathbf{u} -
      \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})\right] \\
    &= \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\left(-\frac{1}{2}\right)(\mathbf{S}^{-1} +
    \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})\dm[\mathbf{u} -
    \mathbf{m}] \\
    &= \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\frac{1}{2}(\mathbf{S}^{-1} + \mathbf{S}^{-\intercal})(\mathbf{u}
    - \mathbf{m}).
  \end{split}
\]
Substituting it back into \eqref{eq:1} gives
\[
  \begin{split}
    \dm\mathbb{E}[V_r(s)] &= \frac{1}{2}\iiint V_r(s)
    (\mathbf{S}^{-1} + \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})
    p(\mathbf{r} | \bm\lambda, \mathbf{X_u}, \mathbf{u}) \mathcal{N}(\mathbf{u};
    \mathbf{m}, \mathbf{S})
    q(\bm\lambda)\,d\mathbf{r}\,d\mathbf{u}\,d\bm\lambda \\
    &= \frac{1}{2}\mathbb{E}[V_r(s) (\mathbf{S}^{-1} +
    \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})].
\end{split}
\]
Hence
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{m}} =
    &-\frac{1}{2}(\mathbb{E}[\Kuu^{-1}] +
    \mathbb{E}[\Kuu^{-1}]^\intercal)\mathbf{m} +
    \mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}] \\
    &- \frac{1}{2}\sum_{i=1}^N\sum_{t=1}^T \mathbb{E}[V_r(s_{i,t})
    (\mathbf{S}^{-1} + \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})] \\
    &- \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')
    \mathbb{E}[V_r(s') (\mathbf{S}^{-1} + \mathbf{S}^{-\intercal})(\mathbf{u} -
    \mathbf{m})].
\end{split}
\]

\subsubsection{$\partial/\partial\mathbf{S}$}

Similarly to the previous section,
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{S}} &=
    \frac{1}{2}\dS\log|\mathbf{S}|
    -\frac{1}{2}\dS\Tr[\mathbb{E}[\Kuu^{-1}]\mathbf{S}] \\
    &- \sum_{i=1}^N \sum_{t=1}^T \dS\mathbb{E}[V_r(s_{i,t})] -
      \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t},
      s')\dS\mathbb{E}[V_r(s')],
  \end{split}
\]
where
\[ \dS\log|\mathbf{S}| = \mathbf{S}^{-\intercal}, \]
and
\[ \dS\Tr[\mathbb{E}[\Kuu^{-1}]\mathbf{S}] = \mathbb{E}[\Kuu^{-1}]^\intercal \]
by \emph{The Matrix Cookbook} \cite{petersen2008matrix}. Then
\[ \dS\mathbb{E}[V_r(s)] = \iiint V_r(s) q(\mathbf{r}) \dS\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})
  q(\bm\lambda)\,d\mathbf{r}\,d\mathbf{u}\,d\bm\lambda, \]
where
\[
  \begin{split}
    \dS\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S}) &=
    \dS\left[\frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\exp \left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right)\right] \\
    &= \dS\left[\frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\right]\exp \left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right) \\
      &+ \frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\dS\left[\exp\left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right)\right] \\
    &=
    \frac{1}{(2\pi)^{m/2}}\dS\left[\frac{1}{|\mathbf{S}|^{1/2}}\right]\exp
    \left( -\frac{1}{2} (\mathbf{u} -
      \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})\right) \\
    &- \frac{1}{2}\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})\dS[(\mathbf{u}
    - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})]. \\
  \end{split}
\]
The two remaining derivatives can be taken with the help of \emph{The Matrix
  Cookbook} \cite{petersen2008matrix}:
\begin{align*}
  \dS\left[\frac{1}{|\mathbf{S}|^{1/2}}\right] =
  -\frac{1}{2}|\mathbf{S}|^{-3/2}\frac{\partial |\mathbf{S}|}{\partial \mathbf{S}} =
  -\frac{1}{2}|\mathbf{S}|^{-3/2}|\mathbf{S}|\mathbf{S}^{-\intercal} &=
  -\frac{1}{2|\mathbf{S}|^{1/2}}\mathbf{S}^{-\intercal}, \\
  \dS[(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
  \mathbf{m})] &= -\mathbf{S}^{-\intercal}(\mathbf{u} - \mathbf{m})(\mathbf{u} -
  \mathbf{m})^\intercal\mathbf{S}^{-\intercal}.
\end{align*}
Plugging them back in gives
\[ \dS\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S}) =
  -\frac{1}{2}\mathbf{S}^{-\intercal}\mathcal{N}(\mathbf{u};
  \mathbf{m}, \mathbf{S}) + \frac{1}{2}\mathcal{N}(\mathbf{u}; \mathbf{m},
  \mathbf{S})\mathbf{S}^{-\intercal}(\mathbf{u} - \mathbf{m})(\mathbf{u} -
  \mathbf{m})^\intercal\mathbf{S}^{-\intercal}, \]
and
\[
  \begin{split}
    \dS\mathbb{E}[V_r(s)] &= \frac{1}{2}\iiint V_r(s)
    (\mathbf{S}^{-\intercal}(\mathbf{u} - \mathbf{m})(\mathbf{u} -
    \mathbf{m})^\intercal\mathbf{S}^{-\intercal} - \mathbf{S}^{-\intercal})
    q(\mathbf{r}) \mathcal{N}(\mathbf{u};
    \mathbf{m}, \mathbf{S}) q(\bm\lambda)\,d\mathbf{r}\,d\mathbf{u}\,d\bm\lambda
    \\
    &= \frac{1}{2}\mathbb{E}[V_r(s)(\mathbf{S}^{-\intercal}(\mathbf{u} -
    \mathbf{m})(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-\intercal} -
    \mathbf{S}^{-\intercal})].
  \end{split}
\]
Therefore
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{S}} &=
    \frac{1}{2}\mathbf{S}^{-\intercal} -
    \frac{1}{2}\mathbb{E}[\Kuu^{-1}]^\intercal - \frac{1}{2}
    \sum_{i=1}^N\sum_{t=1}^T
    \mathbb{E}[V_r(s_{i,t})(\mathbf{S}^{-\intercal}(\mathbf{u} -
    \mathbf{m})(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-\intercal} -
    \mathbf{S}^{-\intercal})] \\
    &- \gamma\sum_{s' \in \mathcal{S}}\mathcal{T}(s_{i,t}, a_{i,t}, s')
    \mathbb{E}[V_r(s')(\mathbf{S}^{-\intercal}(\mathbf{u} -
    \mathbf{m})(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-\intercal} -
    \mathbf{S}^{-\intercal})].
  \end{split}
\]

\subsubsection{$\partial/\partial\alpha_j$}

We begin in the usual way:
    \frac{\partial\mathcal{L}}{\partial\alpha_j} = &-
    \frac{1}{2}\da\mathbb{E}[\Tr[\Kuu^{-2}]] - \frac{1}{2}\da\mathbb{E}[\Tr[\Kuu^{-1}\mathbf{S}]] - \frac{1}{2}\da\mathbb{E}[\mathbf{m}^\intercal\Kuu^{-1}\mathbf{m}] - \frac{1}{2}\da\mathbb{E}[\log|\Kuu|] \\
    &+ \da\mathbb{E}[\mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\mathbf{m}] +
    \da [\alpha_j + \log\Gamma(\alpha_j) + (1 - \alpha_j)\psi(\alpha_j)] \\
      &- \sum_{i=1}^N \sum_{t=1}^T \da\mathbb{E}[V_r(s_{i,t})] -
        \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t},
        s')\da\mathbb{E}[V_r(s')].
First,
\[ \da [\alpha_j + \log\Gamma(\alpha_j) + (1 - \alpha_j)\psi(\alpha_j)] = 1 +
  \psi(\alpha_j) - \psi(\alpha_j) + (1 - \alpha_j)\psi'(\alpha_j) = 1 + (1 -
  \alpha_j)\psi'(\alpha_j) \]
by the definition of $\psi$. The remaining terms can all be treated in the same
way, as they all contain expectations of scalar functions that are independent
of $\alpha_j$, and $\alpha_j$ only occurs in $\Gamma(\lambda_j; \alpha_j,
\beta_j)$. Thus we can work with an abstract function as follows:
    \da\mathbb{E}[f(k_{\bm\lambda}, \mathbf{r})] &= \da \iiint f(k_{\bm\lambda},
    \mathbf{r}) q(\bm\lambda) q(\mathbf{r})
    q(\mathbf{u})\,d\bm\lambda\,d\mathbf{r}\,d\mathbf{u} \\
    &= \!\begin{multlined}[t]
      \iiint f(k_{\bm\lambda}, \mathbf{r}) q(\lambda_0) \cdots q(\lambda_{j-1}) \da
      \left[ \frac{\beta_j^{\alpha_j}}{\Gamma(\alpha_j)} \lambda_j^{\alpha_j -
          1} \right] e^{-\beta_j\lambda_j} \\
      q(\lambda_{j+1}) \cdots q(\lambda_d) q(\mathbf{r})
      q(\mathbf{u})\,d\bm\lambda\,d\mathbf{r}\,d\mathbf{u}.
    \end{multlined}
Then
    \da \left[ \frac{\beta_j^{\alpha_j}}{\Gamma(\alpha_j)} \lambda_j^{\alpha_j - 1} \right] &=
    \frac{\da[\beta_j^{\alpha_j}\lambda_j^{\alpha_j - 1}]\Gamma(\alpha_j) -
      \beta_j^{\alpha_j}\lambda_j^{\alpha_j - 1}\Gamma'(\alpha_j)}{(\Gamma(\alpha_j))^2} \\
    &= \frac{\beta_j^{\alpha_j}\lambda_j^{\alpha_j - 1}\da[\alpha_j\log\beta_j + (\alpha_j -
      1)\log\lambda_j]\Gamma(\alpha_j) - \beta_j^{\alpha_j}\lambda_j^{\alpha_j -
        1}\Gamma'(\alpha_j)}{(\Gamma(\alpha_j))^2} \\
    &= \frac{\beta_j^{\alpha_j}\lambda_j^{\alpha_j - 1}(\log\beta_j + \log\lambda_i)\Gamma(\alpha_j)
      - \beta_j^{\alpha_j}\lambda_j^{\alpha_j - 1}\Gamma'(\alpha_j)}{(\Gamma(\alpha_j))^2} \\
    &= \frac{\beta_j^{\alpha_j}\lambda_j^{\alpha_j - 1}}{\Gamma(\alpha_j)}\left(\log\beta_j +
    \log\lambda_j - \frac{\Gamma'(\alpha_j)}{\Gamma(\alpha_j)}\right),
which means that
    \da\mathbb{E}[f(k_{\bm\lambda}, \mathbf{r})] &= \!\begin{multlined}[t]
      \iiint f(k_{\bm\lambda}, \mathbf{r}) q(\lambda_0) \cdots q(\lambda_{j-1})
      \frac{\beta_j^{\alpha_j}}{\Gamma(\alpha_j)}\lambda_j^{\alpha_j -
        1}e^{-\beta_j\lambda_j} \left(\log\beta_j + \log\lambda_j -
        \frac{\Gamma'(\alpha_j)}{\Gamma(\alpha_j)} \right) \\
      q(\lambda_{j+1}) \cdots q(\lambda_d)
      q(\mathbf{r})q(\mathbf{u})\,d\bm\lambda\,d\mathbf{r}\,d\mathbf{u}
    \end{multlined} \\
    &= \iiint f(k_{\bm\lambda}, \mathbf{r}) \left(\log\beta_j + \log\lambda_j -
      \frac{\Gamma'(\alpha_j)}{\Gamma(\alpha_j)} \right) q(\bm\lambda)
    q(\mathbf{r})q(\mathbf{u})\,d\bm\lambda\,d\mathbf{r}\,d\mathbf{u} \\
    &= \mathbb{E} \left[ f(k_{\bm\lambda}, \mathbf{r}) \left(\log\beta_j +
        \log\lambda_j - \frac{\Gamma'(\alpha_j)}{\Gamma(\alpha_j)} \right)
    \right] \\

>>>>

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\diag}{diag}

@@

\section{Introduction}
%briefly explain the context of the project problem

%Imagine a scenario where the agent is driving on a three-lane highway. The
%features of the environment include the agent's speed, lane number, and
%various distances to different types of vehicles in front of and behind the
%agent. The agent can regulate its speed and position. The main objective is, of
%course, to not crash into other vehicles, but one could also add objectives such
%as staying within the speed limit when close to a police vehicle.

Inverse reinforcement learning (IRL)---a problem proposed by Russell in
1998 \cite{DBLP:conf/colt/Russell98}---asks us to find a reward function for a
Markov decision process that best explains a set of given demonstrations. IRL is
important because reward functions can be hard to define manually
\cite{DBLP:conf/icml/PieterN04,DBLP:journals/corr/abs-1806-06877}, and rewards
are not entirely specific to a given environment, allowing one to reuse the same
reward structure in previously unseen environments
\cite{DBLP:journals/corr/abs-1806-06877,DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
Moreover, IRL has seen a wide array of applications in autonomous vehicle
control \cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} and
learning to predict another agent's behaviour
\cite{DBLP:journals/ai/BogertD18,DBLP:conf/aaai/VogelRGR12,ziebart2008maximum,DBLP:conf/huc/ZiebartMDB08,DBLP:conf/iros/ZiebartRGMPBHDS09}.
Most approaches in the literature (see Section \ref{literature}) make a
convenient yet unjustified assumption that the reward function can be expressed
as a linear combination of features. One proven way to abandon this assumption
is by representing the reward function as a Gaussian process (GP)
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11,DBLP:journals/corr/abs-1208-2112}.
Our main goal is employ variational inference (VI) for recovering the reward
function, which can prove useful in two ways:
\begin{enumerate}
\item As VI tends to be faster than Markov chain Monte Carlo
  sampling \cite{blei2017variational}, we should be able to handle more data.
\item Modelling full posterior distributions for various parameters can result
  in more precise reward predictions, as the model simply holds more
  information.
\end{enumerate}

@@

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood estimates for $\bm\lambda$ and $\mathbf{u}$, and
made an assumption that $\mathbf{r}$ in \eqref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
VI to approximate the full posterior distribution $\posterior$.
\emph{Variational inference} is an approximation technique for probability
densities \cite{blei2017variational}. Let $\approximation$ be our approximating
family of probability distributions for $\posterior$ with its own hyperparameter
vector $\bm\nu$. Then the job of VI algorithms is to optimise $\bm\nu$ in order
to minimise the \emph{Kullback-Leibler} (KL) divergence between the original
probability distribution and our approximation.  KL divergence (asymmetrically)
measures how different the two distributions are, and in this case can be
defined as \cite{blei2017variational}
\[ \begin{split}
    \DKL(\approximation || \posterior) &= \Eq [\log\approximation -
    \log\posterior ] \\
    &= \Eq[\log\approximation - \log\pfull] \\
    &+ \Eq[\log p(\mathcal{D}, \mathbf{X_u})].
  \end{split}
\]
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as Lebesgue integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L}(\bm\nu) &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\bm\lambda\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\end{equation}
% TODO: are they actually Lebesgue? What about notation?

By considering full probability distributions instead of point estimates---as
long as the approximations are able to capture important features of the
posterior---our predictions are likely to be more accurate and rely on fewer
assumptions. Moreover, we hope to make use of various recent advancements in VI
for both time complexity and approximation distribution fit (see Section
\ref{literature}), making the resulting algorithm competitive both in terms of
running time and model fit. Therefore, we raise the following research
questions\footnote{For each question, we are primarily comparing the original GP
IRL approach \cite{DBLP:conf/nips/LevinePK11} with our variational version.}:
\begin{enumerate}
\item Does VI allow us to make more accurate reward predictions when given the
  same amount of training time?
\item How does the reward prediction accuracy differ between VI and sampling
  approaches when given the same amount of training data?
\item How well does our VI model approximate the real posterior distribution?
\item Is VI better at learning behaviour driven by multiple goals?
\item Is VI more robust at handling mistakes in demonstrations?
\end{enumerate}

@@

Finally, instead of using GPs to model nonlinear reward functions, one can use a
neural network (NN), as demonstrated by Wulfmeier et al.
\cite{wulfmeier2015maximum}. Their approach benefits from constant time
inference and the ability to learn complex features either from already-given
features or even from raw data. The only disadvantage (as demonstrated in the
paper) is that NNs take longer to learn compared to GPs.

@@

Variational inference has seen a recent increase in interest among academics,
with different approaches focusing on different goals: better time complexity,
handling a wider variety of models, making approximations more accurate, and
using more complex function approximation techniques (such as neural networks)
to infer local latent variable values without having to calculate them
individually for each data point \cite{DBLP:journals/corr/abs-1711-05597}. As
our IRL model is based on a GP, we will begin by reviewing some of the VI
approaches applied specifically to GP regression. Based on a recent review of
scalable GPs \cite{DBLP:journals/corr/abs-1711-05597}, we will concentrate on
\emph{stochastic variational sparse approximations}, as they have achieved
modelling accuracy close to that of the full GP with no approximations, with
many methods providing a time complexity of $\mathcal{O}(m^3)$. Below we provide
a short overview of various assumptions that have been used in approximating
\emph{sparse} GPs (i.e., GPs that use inducing points), following on a paper by
Qui\~{n}onero-Candela and Rasmussen \cite{DBLP:journals/jmlr/CandelaR05}.

@@

Table \ref{table:approximations} further summarises some of the recent and/or
influential GP approximation approaches that might be suitable for our GP IRL
model. In order to derive a reliable ELBO, the inducing points should be either
fixed or modelled by a probability distribution (none of the approaches do
that), while hyperparameters cannot be variational (the reason will become clear
in Section \ref{sec:proposed_approach}). We would also like to avoid having the
hyperparameters fixed, as learning them efficiently would introduce a separate
problem. As the approach by Hoang et al. does not seem to be easily extendable
to modelled hyperparameters, it is unsuitable to our needs.

The variational Fourier features (VFF) algorithm by Hensman et al.
\cite{DBLP:journals/jmlr/HensmanDS17} is only defined for Mat\'ern kernels,
which is likely to be too restrictive for our situation (the original paper on
using GPs for IRL \cite{DBLP:conf/nips/LevinePK11} used the automatic relevance
detection kernel that has weights controlling how important each feature is).
While extending VFF to support a flexible class of kernels defined by Wilson and
Adams \cite{pmlr-v28-wilson13} is an interesting and promising avenue of work,
it is likely to be beyond the scope of this project.

The derivation of the ELBO in the work of Peng et al.
\cite{DBLP:conf/icml/PengZZQ17} relies primarily on the fact that the evidence
for a GP is a Gaussian, while in our case the evidence can be defined in several
ways depending on the model, but must always involve $\mathcal{D}$, making the
main idea of the paper inapplicable to IRL. Remaining papers can be rejected for
similar reasons, as they are all highly dependent on the specific situation of
approximating a GP regression. Thus we cannot simply apply a GP VI model as a
whole, although there might be some smaller ideas that we can use (some of which
will be mentioned in the next section).

Finally, since we expect $\posterior$ to be highly irregular, we would like our
approximation to be capable of representing a wide range of possible probability
distributions. The primary way of representing complex posteriors in VI is by
using \emph{normalising flows}, i.e., a collection of invertible
functions---parametrised by additional variational parameters---that are applied
to latent variables \cite{DBLP:conf/icml/RezendeM15}. Unfortunately, this
parametrisation also means that the gradient of the joint probability
distribution w.r.t. variational parameters $\nabla_{\bm\nu}\pfull$ is no longer
zero, making an analytic expression for the ELBO impossible using the usual
methods. While it may be possible to have an approximating distribution for the
flow parameters, it is uncertain how such an algorithm would behave, as it would
have to perform optimisation in a space with significantly more dimensions.

@@

\subsection{Structure of the Approximating Distribution}

@@

\subsection{Variational Inference} \label{vi_algs}

The typical way to optimise a quantity (the ELBO, in this case) involves
computing its gradient \cite{blei2017variational}. Unfortunately, some of the
terms in $\mathcal{L}$ are still left as expected values. \emph{Black box
  variational inference} (BBVI) \cite{DBLP:conf/aistats/RanganathGB14} suggests
a way to express the gradient as an expectation without having to take the
gradient of the posterior:
\[ \nabla_{\bm\nu} \mathcal{L} = \Eq[\nabla_{\bm\nu} \log \approximation
  (\log \pfull - \log \approximation)]. \]
The gradient of the ELBO then has a unbiased estimate
\[ \nabla_{\bm\nu} \mathcal{L} \approx \frac{1}{S} \sum_{s=1}^S
  \nabla_{\bm\nu} \log \approximationS (\log \pfullS - \log
  \approximationS) \]
computed by drawing $S$ Monte Carlo samples $(\bm\lambda_s, \mathbf{u}_s,
\mathbf{r}_s) \sim \approximation$.

The main theorem behind the BBVI trick is the Lebesgue's dominated convergence
theorem \cite{royden1988real}, which---under some conditions---allows one to
claim that:
\[ \lim_{n \to \infty} \int_E f_n = \int_E \lim_{n \to \infty} f_n. \]
We can use the same idea on parts of $\mathcal{L}$ that we cannot take the
derivative of otherwise. In this case, we want a gradient instead of a limit,
and the integral represents an expected value.
